---
title: "RepData_PeerAssessment1"
author: "joedinoto"
date: "7/6/2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Load the data into a dataframe.
```{r }
unzip("activity.zip")
DF<- read.csv("activity.csv")
```
Invoke the powers of the tidyverse to make the date column read as dates, the interval column as factors, and turn the whole dataframe into a tibble.
```{r}
library(tidyverse)
library(lubridate)
DF$date <- ymd(DF$date)
DF$interval <- as.factor(DF$interval)
DF<- as_tibble(DF)
```


## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

 1. Make a histogram of the total number of steps taken each day
 
First, figure out the total number of steps taken each day using dplyr. 
```{r}
newDF <- na.omit(DF)
by_date<- newDF %>%
  group_by(date,add=TRUE) %>%
  summarise(total_steps=sum(steps))
by_date
```

Now create the histogram.
```{r}
with(by_date,hist(total_steps))
```

 2. Calculate and report the mean and median total number of steps taken per day
```{r}
by_date_mean <- newDF %>%
  group_by(date) %>%
  summarise(mean(steps))
by_date_mean
```